Income Disparity and Its Influence on Higher Education Affordability and Access

### 2024-02-19 starter_folder-main 4/paper/paper.qmd
updated tables/abstract/limitations

---
title: "Income Disparity and Its Influence on Higher Education Affordability and Access"
author: 
  - Michael Fang
  - Harrison Huang
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

1. Recreating the Model: We will start by reproducing the competitive college market model used in the original study, ensuring our work is entirely reproducible. This step includes verifying the model's assumptions, computational methods, and data inputs.
2. Expanding the Study with an analysis on Canadian Universities: Specifically examine how income inequality within Canada affects university tuition, comparing these dynamics with the findings from the U.S. context. This will provide insights into how different policy environments and socio-economic conditions influence the relationship between income inequality and higher education costs.

@@

@@

@@

```
\newpage

```{r}
#| message: false
#| echo: false
#| label: tbl-public_data
#| tbl-cap: Public Four Year College Data

# create frame for public data
public_data <- tibble("Type" = expenditure_type, "Data_1" = pub_earliest_data, "Data_2" = pub_latest_data, "Percent_Change" = pub_percent_change, "Time_Frame" = pub_year_frame)

```{r}
#| message: false
#| echo: false
#| label: tbl-private_data
#| tbl-cap: Private Four Year College Data

# create frame for private data
private_data <- data.frame("Type" = expenditure_type, "Data_1" = pri_earliest_data, "Data_2" = pri_latest_data, "Percent_Change" = pri_percent_change, "Year_Frame" = pri_year_frame)

@@

@@

\appendix

# Appendix {-}

# Additional data details

>>>>

---
title: "Income Disparity and Its Influence on Higher Education Affordability and Access"
author: 
  - Michael Fang
  - Harrison Huang
thanks: "Code and data are available at: https://github.com/fanger2791/College-Tuition-and-Income-Inequality/tree/main"
date: today
date-format: long
abstract:   Throughout the years, we have seen how rising income inequality fuels college tuition hikes, casting a spotlight on the urgent need for policies that bridge the educational divide. This paper is a reproduction on the connection between rising income inequality and the surge in college tuition costs in the United States. Utilizing a detailed economic model, the increasing disparity in income has been a significant driver behind the growth in tuition fees since 1990, accounting for more than half of the observed increase and has also led to a decline in college attendance rates. Extending this analysis through secondary research to include Canadian universities reveals similar trends, indicating a broader North American pattern where income inequality impacts educational accessibility. This comparative approach underscores the global relevance of their findings, suggesting that addressing income disparities could play a crucial role in making higher education more accessible.
format: pdf
number-sections: true
bibliography: references.bib
---

@@

1. Recreating Table 3: We will split the table into private college versus public college. In the original work, it was not clearly stated where the data came from or for private two year studies or public four year studies. With that, we will look at data sets from the paper and recreate Table 3 from the original work. The new tables will focus on the price change from the earliest data and most recent data. In addition, the two tables are separated by private four year non profit college versus public four year college. This step includes extensive data cleaning and edits due to excel file formats. All the steps can be found in the source code to increase the reproducibility of our analysis.

2.  Recreating the Model: We will start by reproducing the competitive college market model used in the original study, ensuring our work is entirely reproducible. This step includes verifying the model's assumptions, computational methods, and data inputs.

3.  Expanding the Study with a Canadian Lens: Specifically examine how income inequality within Canada affects university tuition, comparing these dynamics with the findings from the U.S. context. This will provide insights into how different policy environments and socio-economic conditions influence the relationship between income inequality and higher education costs.

@@

R was the language and environment used for the bulk of this reproduction, alongside

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-public_data
#| tbl-cap: Public Four Year College Data

# create frame for public data
public_data <- tibble("Type" = expenditure_type, "Data_1" = pub_earliest_data, "Data_2" = pub_latest_data, "PercentChange" = pub_percent_change, "TimeFrame" = pub_year_frame)

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-private_data
#| tbl-cap: Private Four Year College Data

# create frame for private data
private_data <- data.frame("Type" = expenditure_type, "Data_1" = pri_earliest_data, "Data_2" = pri_latest_data, "PercentChange" = pri_percent_change, "YearFrame" = pri_year_frame)

In the two tables @tbl-public_data and @tbl-private_data, the first column stands for tuition, fees, room and board. This is a summary data of the entire expense at that specific institution. The Grants by Income data then looks grants by year given to each income group. For example, the second row of the table looks at grants given out to people that has an annual income of less than 30,000. The sixth row looks at state funding according to an individual's income by per 1000. For example, if an individual were to make 50,000 a year, state funding for higher education to that individual should be 368.5$. There was no private and public data, as a result the state funding data is the same for both private and public colleges. The last two row looks at the data of the institution's revenue and expenditure on subsidizing students. The year frame was different for each data sets, therefore, we decided to create a separate column for cleaner look for the table. This would prevent N/A cells and could be confusing for readers.

First, looking at @tbl-public_data, there is a big increase in the net fees at 62.77% increase in just 20 years. Then looking at the grants given out by the public institutions, even though we see a continuous increase of more funding for the two lower income groups, we are still seeing a big increase in percentage for higher income individuals. When looking at state funding, there is even a -28.31 % decrease in the funding for higher education. Lastly looking at the tuition revenue and subsidy, despite making steady profits, the institution on the other hand decreased subsitizing expenditures by 8.5% from 2003 to 2014.

Second, here are some key findings from @tbl-private_data. There is a less increase of fees at 30.27%. However, the total expenditure is significant more compared to public institutions. Then looking at the grants given out by private institutions, with the lowest income group gaining a significant increase of 120%. Interestingly, the highest group also gained 97.36%. In the same situation, state funding, showed a -28.31 % decrease for higher education. Lastly looking at the tuition revenue and subsidy, 17.62% increase in tuition revenue, and also an increase of 28.86% subsidy expenditures.

Comparing the two data set, we can see that even though the percentage increase of private institution fees. The net expenditure is still almost three times of public institutions. Overall by income group, private institutions are also a lot more generous in giving out grants, with large numbers compared to public institutions. Private institutions also have a positive percentage increase of subsidy expenditure compared to public institutions despite both having positive tuition revenue over the years. From the table, we can see that private institutions are much more generous in terms of grants and expenditure studies, however, they are also a lot more expensive to attend compared to public schools.

In attempts to uncover more about income disparity and the accesses of higher education, we will recreate two models from the paper.

@@

@@

There are several potential weaknesses in the findings that warrant consideration. Firstly, the model's assumptions, including a continuous distribution of college quality and linear relationships between variables, might oversimplify the complex, non-linear dynamics present in real-world scenarios. This simplification could introduce inaccuracies in predicting tuition trends and attendance patterns. Secondly, the study's reliance on observed data trends and specific calibrations may not fully encapsulate the intricacies of income distribution, tuition policies, and other socioeconomic factors, leading to potential biases or limitations in the dataset used for analysis.

Moreover, while the model accounts for a broad spectrum of college quality, it might not accurately represent specific types of educational institutions, such as community colleges, vocational schools, and online universities, which do not adhere to the traditional metrics of college quality. Additionally, the focus on income inequality as the primary driver overlooks other significant factors, such as changes in government funding for education, demographic shifts, and evolving labor market demands, which could also influence tuition costs and attendance rates.

Furthermore, the model may not sufficiently consider how policy changes or shifts in household behavior in response to rising tuition and income inequality could alter the college market dynamics. For instance, the increasing prevalence of online education and alternative credentialing pathways might disrupt traditional patterns of college attendance and tuition pricing strategies. Lastly, the study's conclusions are based on historical data up to a certain point, and the long-term applicability of these findings might be limited by emerging trends and changes in the landscape of higher education and economic inequality.

## Limitations to reproducing

From a reproducibility standpoint, the absence of provided code or detailed methodology for recreating the tables, graphs, and statistical analyses by Zhifeng Cai and Jonathan Heathcote introduces significant limitations. The ability to replicate study findings is foundational to the validation and advancement of scientific research. Without access to the exact computational procedures, data analysis scripts, or software used, other researchers are hindered in their ability to verify the results, explore the implications further, or apply the methodology to new datasets. This lack of transparency can impede the reproducing process as others may struggle to build directly on these findings without a clear understanding of the original analysis techniques.

Moreover, the reproducibility issue extends beyond just the recreation of visualizations to the very core of the study's conclusions. The modeling assumptions, data transformations, and statistical methods employed are crucial for understanding the dynamics between income inequality and college tuition. Without clear documentation or code, it becomes challenging to assess the robustness of the model, the sensitivity of the results to different assumptions, or the potential for bias in the analysis. This gap not only affects the credibility of the findings but also limits the potential for educational policymakers, economists, and other stakeholders to apply these insights effectively.

@@
